---
title: "Getting Started with geefetch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with geefetch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

**geefetch** provides a unified R interface for extracting environmental
covariates from Google Earth Engine (GEE). Unlike other GEE R packages,
geefetch works **without Python** — it uses the GEE REST API directly via
`httr2` and `gargle`.

Key features:

- **No Python required** — installs like any normal R package
- **19+ built-in datasets** — vegetation, climate, soil, topography
- **Single-function extraction** — `read_gee()` dispatcher + convenience aliases
- **Batch extraction** — `collect_gee_data()` for multi-location, multi-date, multi-dataset
- **Automatic caching** — repeated queries are served from disk
- **User-extensible** — register any GEE collection with `gee_register_dataset()`

## Installation

```{r install, eval = FALSE}
# From GitHub (development version, with vignettes)
remotes::install_github("AAGI-AUS/geefetch", subdir = "geefetch",
                        build_vignettes = TRUE)
```

## Authentication

geefetch uses the same Google OAuth flow as `googlesheets4`, `googledrive`,
and `bigrquery` via the `gargle` package.

```{r auth, eval = FALSE}
library(geefetch)

# Interactive OAuth (opens browser)
gee_auth()

# Check your connection
gee_status()
```

```
## -- geefetch status --
## v Authenticated: yes
## i Project: "earthengine-legacy"
## i Backend: REST API (httr2 + gargle)
## i Cache dir: '~/Library/Caches/.../geefetch'
## i Registered datasets: 19
```

For non-interactive use (CI, HPC), use a service account:

```{r auth-sa, eval = FALSE}
gee_auth(path = "path/to/service-account.json")
```

For first-time setup, `gee_setup()` provides step-by-step instructions.

## Your first extraction

Extract MODIS NDVI for a region in South Australia:

```{r first-extraction, eval = FALSE}
library(terra)

# Define a region of interest
aoi <- ext(138, 140, -36, -34)

# Extract NDVI for a specific date
ndvi <- read_modis_ndvi(date = "2024-06-15", region = aoi)

# Plot the result
plot(ndvi, main = "MODIS NDVI — 15 June 2024")
```

The result is a standard `terra::SpatRaster` — you can use all `terra`
functions for analysis, cropping, masking, and export.

## Point extraction

To extract values at specific coordinates, use `terra::extract()`:

```{r point-extraction, eval = FALSE}
# Define some sites
sites <- data.frame(
  lon = c(138.6, 139.5, 140.2),
  lat = c(-34.9, -35.5, -34.2)
)

# Convert to spatial points
pts <- vect(sites, geom = c("lon", "lat"), crs = "EPSG:4326")

# Extract NDVI at the sites
values <- extract(ndvi, pts)
print(values)
```

```
##   ID      NDVI
## 1  1 0.3245
## 2  2 0.4512
## 3  3 0.2891
```

## Using the dispatcher

All convenience functions (`read_modis_ndvi()`, `read_era5()`, etc.) are
thin wrappers around the central `read_gee()` dispatcher. You can use
either form:

```{r dispatcher, eval = FALSE}
# These are equivalent:
ndvi <- read_modis_ndvi(date = "2024-06-15", region = aoi)
ndvi <- read_gee("MODIS_NDVI", date = "2024-06-15", region = aoi)
ndvi <- read_gee("NDVI", date = "2024-06-15", region = aoi)  # alias
```

## Browsing available datasets

```{r datasets, eval = FALSE}
gee_datasets()
```

```
##            dataset                          collection       domain resolution temporal
##  1:     modis_ndvi                  MODIS/061/MOD13A2   Vegetation     1000m    16day
##  2:      modis_lst                  MODIS/061/MOD11A2  Temperature     1000m     8day
##  3:      era5_temp       ECMWF/ERA5_LAND/DAILY_AGGR      Climate    11132m    daily
##  ...
## 19: openlandmap_ph  OpenLandMap/SOL/SOL_PH-H2O_...  Soil (Global)      250m   static
```

## Caching

geefetch automatically caches extraction results to disk. Repeated
identical queries are served instantly:

```{r caching, eval = FALSE}
# First call: hits GEE REST API (~2-5 seconds)
elev <- read_srtm(region = aoi)

# Second call: served from disk cache (< 10 ms)
elev <- read_srtm(region = aoi)

# Check cache size
gee_status()

# Clear cache if needed
gee_clear_cache()
```

## Next steps

- See `vignette("collect_gee_data")` for batch extraction across multiple
  locations, dates, and datasets
- See `vignette("custom_datasets")` for registering your own GEE collections
- See `gee_datasets()` for the full list of supported datasets
