---
title: "Getting Started with geefetch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with geefetch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

**geefetch** provides a unified R interface for extracting environmental
covariates from Google Earth Engine (GEE). Unlike other GEE R packages,
geefetch works **without Python** — it uses the GEE REST API directly via
`httr2` and `gargle`.

Key features:

- **No Python required** — installs like any normal R package
- **19+ built-in datasets** — vegetation, climate, soil, topography
- **Single-function extraction** — `read_gee()` dispatcher + convenience aliases
- **Batch extraction** — `collect_gee_data()` for multi-location, multi-date, multi-dataset
- **Automatic caching** — repeated queries are served from disk
- **User-extensible** — register any GEE collection with `gee_register_dataset()`

## Installation

```{r install, eval = FALSE}
# From R-Universe (binaries, updated ~1 h after merge to main):
options(repos = c(
  "aagi-aus" = "https://aagi-aus.r-universe.dev",
  CRAN       = "https://cran.r-project.org"
))
install.packages("geefetch")

# Development version (latest commit on main):
pak::pak("AAGI-AUS/geefetch", subdir = "geefetch")
```

## Authentication

**Read this section fully before running `gee_auth()`.** Skipping ahead is
the fastest way to spend an afternoon debugging HTTP 403s.

Authentication to Google Earth Engine has two independent concepts that
`geefetch` exposes together:

| Concept | What it is | Where it lives |
|---|---|---|
| **Identity** | Which Google account you're signing in as | OAuth token |
| **Resource project** | Which Google Cloud project owns your EE calls, quotas, and billing | `project = ` argument to `gee_auth()` and `X-Goog-User-Project` header |

If the identity is signed into account A but the resource project was
created under account B, you'll get HTTP 403. The rest of this section is
about making both match.

### Step 1 — Create a Google Cloud project and register it with Earth Engine

You only do this once, before you ever touch R.

1. **Choose one Google account.** The rest of this setup — the Cloud
   project, the EE registration, the OAuth consent — must all happen under
   the *same* Google account. Using your institutional account
   (`*.edu.au`, `*.gov`, etc.) is typical for academic / research work.
   A personal Gmail account works too. Decide now and stick with it.

2. **Create a dedicated Google Cloud project** at
   <https://console.cloud.google.com/projectcreate>. Give it a memorable
   name (e.g. `my-geefetch-dev`). Google will assign a short **Project ID**
   (the string you type, e.g. `my-geefetch-dev`) and a long **Project
   number** (a 12-digit integer, e.g. `123456789012`). Copy both. You'll
   use the **project number** in R — it's unambiguous across accounts.

3. **Register the project for Earth Engine noncommercial use** at
   <https://code.earthengine.google.com/register>. Choose *Academic* or
   *Research*, select your new project from the dropdown, fill in the
   institution details, submit. Academic approval is usually instant.

4. **Enable the Earth Engine API** on the project at
   <https://console.cloud.google.com/apis/library/earthengine.googleapis.com>
   — make sure the project picker top-left shows your project, then click
   **Enable**.

5. **Verify** at <https://console.cloud.google.com/apis/dashboard> that
   "Earth Engine API" appears in the enabled APIs list for your project.

### Step 2 — Authenticate in R

Use the **project number**, not the project ID string. This is the single
most common cause of the 403 "Earth Engine API has not been used"
error — passing the ID can silently resolve to a different project
number under a different identity.

```{r auth, eval = FALSE}
library(geefetch)

gee_auth(
  project = "123456789012",              # <-- YOUR project number as a string
  email   = "you@gmail.com"              # <-- the exact Google account
)
```

A browser window will open for Google OAuth (same flow as `googlesheets4`).
Sign in as the account matching `email`, grant consent for Earth Engine
and Cloud Platform scopes, close the tab.

### Step 3 — Verify

```{r verify, eval = FALSE}
gee_status()
```

```
## -- geefetch status --
## v Authenticated: yes
## i Project: "123456789012"
## i Backend: REST API (httr2 + gargle)
## i Cache dir: '~/Library/Caches/.../geefetch'
## i Registered datasets: 19
```

The `Project:` line must show your project **number**. If it shows an
ID string like `"earthengine-legacy"` or a different number, stop and
re-run `gee_auth()` with the correct arguments.

### Step 4 — Persist across sessions

Add to your `.Rprofile` (or the top of each analysis script) so you don't
need to re-type the project number:

```{r rprofile, eval = FALSE}
options(geefetch.project = "123456789012")
# Then gee_auth() picks it up without explicit project arg.
```

### Non-interactive authentication

For CI, HPC, or any unattended context, use a service-account key file.
Create the key on the same Cloud project used above
(<https://console.cloud.google.com/iam-admin/serviceaccounts>), grant it
the **Earth Engine Resource Writer** role, download the JSON, then:

```{r auth-sa, eval = FALSE}
gee_auth(
  project = "123456789012",
  path    = "path/to/service-account.json"
)
```

Store the key file outside your repository. Never commit it.

### Troubleshooting — HTTP 403 decision tree

If `read_modis_ndvi()` or any `read_*()` call returns a 403 after
authentication appeared to succeed, the error message contains a project
number. Match that number against what you expect:

| Error's project number | Meaning | Fix |
|---|---|---|
| Matches your `project = ` | Earth Engine API not actually enabled, or registration not yet propagated | Re-verify Step 1 items 3 + 4; wait 2-3 minutes and retry |
| **Different** number | OAuth token's quota project disagrees with your resource project | Restart R; wipe gargle's token cache (`unlink(list.files(gargle::gargle_oauth_cache(), full.names = TRUE))`); re-run `gee_auth()` with the project **number**, not the ID |

If neither helps, `gee_setup()` prints the full check-list and
links to the three Google Cloud Console pages involved.

## Your first extraction

Extract MODIS NDVI for a region in South Australia:

```{r first-extraction, eval = FALSE}
library(terra)

# Define a region of interest
aoi <- ext(138, 140, -36, -34)

# Extract NDVI for a specific date
ndvi <- read_modis_ndvi(date = "2024-06-15", region = aoi)

# Plot the result
plot(ndvi, main = "MODIS NDVI — 15 June 2024")
```

The result is a standard `terra::SpatRaster` — you can use all `terra`
functions for analysis, cropping, masking, and export.

## Point extraction

To extract values at specific coordinates, use `terra::extract()`:

```{r point-extraction, eval = FALSE}
# Define some sites
sites <- data.frame(
  lon = c(138.6, 139.5, 140.2),
  lat = c(-34.9, -35.5, -34.2)
)

# Convert to spatial points
pts <- vect(sites, geom = c("lon", "lat"), crs = "EPSG:4326")

# Extract NDVI at the sites
values <- extract(ndvi, pts)
print(values)
```

```
##   ID      NDVI
## 1  1 0.3245
## 2  2 0.4512
## 3  3 0.2891
```

## Using the dispatcher

All convenience functions (`read_modis_ndvi()`, `read_era5()`, etc.) are
thin wrappers around the central `read_gee()` dispatcher. You can use
either form:

```{r dispatcher, eval = FALSE}
# These are equivalent:
ndvi <- read_modis_ndvi(date = "2024-06-15", region = aoi)
ndvi <- read_gee("MODIS_NDVI", date = "2024-06-15", region = aoi)
ndvi <- read_gee("NDVI", date = "2024-06-15", region = aoi)  # alias
```

## Browsing available datasets

```{r datasets, eval = FALSE}
gee_datasets()
```

```
##            dataset                          collection       domain resolution temporal
##  1:     modis_ndvi                  MODIS/061/MOD13A2   Vegetation     1000m    16day
##  2:      modis_lst                  MODIS/061/MOD11A2  Temperature     1000m     8day
##  3:      era5_temp       ECMWF/ERA5_LAND/DAILY_AGGR      Climate    11132m    daily
##  ...
## 19: openlandmap_ph  OpenLandMap/SOL/SOL_PH-H2O_...  Soil (Global)      250m   static
```

## Caching

geefetch automatically caches extraction results to disk. Repeated
identical queries are served instantly:

```{r caching, eval = FALSE}
# First call: hits GEE REST API (~2-5 seconds)
elev <- read_srtm(region = aoi)

# Second call: served from disk cache (< 10 ms)
elev <- read_srtm(region = aoi)

# Check cache size
gee_status()

# Clear cache if needed
gee_clear_cache()
```

## Next steps

- See `vignette("collect_gee_data")` for batch extraction across multiple
  locations, dates, and datasets
- See `vignette("custom_datasets")` for registering your own GEE collections
- See `gee_datasets()` for the full list of supported datasets
